---
title: "Week 2 Course Project 1"
author: "Keith Webster"
date: "May 29, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reproducible Research Week 2 Course Project 1
## Load required libraries and process activity data
### Unzip archive file and read data into data frame
```{r}
# Graphics library
library(ggplot2)

# Unzip zip file
unzip("repdata%2Fdata%2Factivity.zip")

# Read activity data into data frame "activity_data"
activity_data <- read.csv("activity.csv")
```
### Summarize Activity Data
```{r}
# Display head activity data
head(activity_data)
```
```{r}
### Data dimensions of activity_data
dim(activity_data)
```
```{r}
### Data summary for activity_data
summary(activity_data)
```
### NA and Date String Modification
```{r}
## Data processing NA and date formatting

### Generate NA Vector
activity_data_NA <- is.na(activity_data$steps)
### Remove NA data from file and store new file
activity_data_nona <- activity_data[!activity_data_NA,]
### Convert date type to date class
activity_data$date <- as.Date(as.character(activity_data$date))
```
## What is the mean total number of steps take per day?
### 1. Calculate the total number of steps taken per day
```{r}
# aggregate clean non NA steps per day (SUM)
daily_aggregate <- aggregate(activity_data$steps, by=list(activity_data$date), sum)
# adjust column names
names(daily_aggregate)[1] = "date"
names(daily_aggregate)[2] = "totalsteps"
# top 15 of summed steps by day
head(daily_aggregate,15)
```
### 2. Make a histogram of the total number of steps taken each day
```{r}
# Plot using ggplot
ggplot(daily_aggregate, aes(x = totalsteps)) +
  geom_histogram(fill = "steelblue", binwidth=1000) +
  labs(title = "Total Daily Steps", x = "Steps", y = "Frequency")

```


### 3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
# Mean of steps taken per day
mean(daily_aggregate$totalsteps,na.rm=TRUE)
```

```{r}
# Mean of steps taken per day
median(daily_aggregate$totalsteps,na.rm=TRUE)
```

## What is the average daily activity pattern?
### 1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
Mean_Data_ByInterval <- aggregate(activity_data_nona$steps, by=list(activity_data_nona$interval), mean)

#set the column names
names(Mean_Data_ByInterval)[1] = "interval"
names(Mean_Data_ByInterval)[2] = "steps"

ggplot(Mean_Data_ByInterval, aes(x = interval, y = steps)) +
  labs(title = "Sum of Steps by Interval", x = "interval", y = "steps")+
  geom_line(color="steelblue")
```

### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum numbers of steps?

```{r}
max_interval <- Mean_Data_ByInterval[which.max(Mean_Data_ByInterval$steps),]
max_interval
```

## Impute missing values from activity data set
### 1. Calculate and report the total number of missing values in the dataset

```{r}
# Generate list of NAs
missing_values <- sum(activity_data_NA)
missing_values
```

### 2. Use mean interval steps for filling in all of the missing values in the activity dataset
### 3. Create a new dataset that replaces the na data in the original activity data set

```{r}
# Generate enhanced activity data
activity_data_2 <- activity_data
# NA's in the activity dataset
# dataset minus NAs for mean calc
NA_activity_data_2 <- activity_data_2[is.na(activity_data_2$steps),]
clean_activity_data_2 <- activity_data_2[!is.na(activity_data_2$steps),]
```
```{r}
# Generate mean data_enhanced by interval
mean_data2_byinterval <- aggregate(clean_activity_data_2$steps, by=list(clean_activity_data_2$interval), sum)
names(mean_data2_byinterval)[1] = "interval"
names(mean_data2_byinterval)[2] = "steps"
# Impute values
activity_data_2 <- activity_data
missingData <- is.na(activity_data_2$steps)
meanVals <- tapply(activity_data_nona$steps, activity_data_nona$interval, mean, na.rm=TRUE, simplify=TRUE)
activity_data_2$steps[missingData] <- meanVals[as.character(activity_data_2$interval[missingData])]
```
```{r}
# sum missing data count
sum(missingData)
# count of NA values
sum(is.na(activity_data_2$steps))
```
### 4.1 Make a histogram of the total steps taken each day and Calculate and report the mean and median total number of steps taken per day.

```{r}
FullSummedDataByDay <- aggregate(activity_data_2$steps, by=list(activity_data_2$date), sum)

names(FullSummedDataByDay)[1] ="date"
names(FullSummedDataByDay)[2] ="totalsteps"
head(FullSummedDataByDay,15)
```
```{r}
#Plot using ggplot
ggplot(FullSummedDataByDay, aes(x = totalsteps)) +
  geom_histogram(fill = "steelblue", binwidth=1000) +
  labs(title = "Total Daily Steps", x = "Steps", y = "Frequency")
```
```{r}
# Mean on New Data
mean(FullSummedDataByDay$totalsteps)
```
```{r}
#Median on New Data
median(FullSummedDataByDay$totalsteps)
```
### 4.2 Do these values differ from the estimates in the first part of the assignment?

### Yes the mean is the same but the median has increased.

### 4.3 What is the impact of imputing the missing data on the estimates of the total number of daily steps?

### Use of mean data as the data input method skews data towards the mean

## Are there differences in activity patterns between weekdays and weekends?
```{r}
activity_data_2$weekday <- weekdays(activity_data_2$date)
activity_data_2$weekend <- ifelse (activity_data_2$weekday == "Saturday" | activity_data_2$weekday == "Sunday", "Weekend", "Weekday")
#activity_data_2$weekend <- as.factor(activity_data_2$weekend)
head(activity_data_2,5)
```
```{r}
MeanDataWeekendWeekday <- aggregate(activity_data_2$steps, by=list(activity_data_2$weekend, activity_data_2$interval), mean)
names(MeanDataWeekendWeekday)[1] ="weekend"
names(MeanDataWeekendWeekday)[2] ="interval"
names(MeanDataWeekendWeekday)[3] ="steps"

ggplot(MeanDataWeekendWeekday, aes(x = interval, y=steps, color=weekend)) +
  geom_line() +
  facet_grid(weekend ~ .) +
  labs(title = "Mean of Steps by Interval", x = "interval", y = "steps")
```

### There appears to be higher steps per day on the weekend implying more physical activity which makes sense.



  


  

